---
title: "Implementing Cure Threshold"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Implementing Cure Threshold}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

hooks = knitr::knit_hooks$get()
hook_foldable = function(type) {
  force(type)
  function(x, options) {
    res = hooks[[type]](x, options)
    
    if (!isTRUE(options[[paste0("fold.", type)]])) {
      # By default, set folding to FALSE unless explicitly set to TRUE
      
      return(res) 
    } else {
      # Use the chunk label as the summary if it exists, otherwise fallback to type
      summary_title <- options$fold.title
      type <- ifelse(type == "source", "SourceCode", type)
      if (is.null(summary_title)) {
        summary_title <- type
      } else {
        summary_title <- paste0(summary_title, " - ", type)
      }
      
      return(
        paste0(
          "<details><summary>", summary_title, "</summary>\n\n",
          res,
          "\n\n</details>"
        )
      )
    }
  }
}

knitr::knit_hooks$set(
  output = hook_foldable("output"),
  plot = hook_foldable("plot"),
  source = hook_foldable("source")
)
```

## Introduction:

The Neuroblastoma Partitioned Survival Model (PSM) employs a *cure threshold*. The cure threshold is the time in years at which the model survival assumptions change, and is set to `r NeuroblastomaPSM::l_psm_parameters$cure_threshold` years in the base-case analysis. Beyond the cure threshold, the EFS curve only represents death events, as it was assumed that patients who were Event-Free at this point were cured. The mortality hazard rates for Event-Free and Relapsed patients are set to `r NeuroblastomaPSM::l_psm_parameters$smr_EFS` and `r NeuroblastomaPSM::l_psm_parameters$smr_PPS` times the mortality hazard rates for the general population, respectively.

This vignette describes how the PSM:

- adjusts general population hazard rates, and
- employs the cure threshold assumption.

## Ajusting general population hazard rates:

General population hazard rates can be found in Life Tables. The life table of Jordan is shipped with the `NeuroblastomaPSM` package, and is used by default in the PSM model. The following steps describe how the `NeuroblastomaPSM` package adjusts the baseline mortality.

### Generating the general population survival curve:

The survival curve represents the probability of surviving up to the corresponding age or time point. The `get_lifeTable_survival_curve` function calculates the survival curve from a given vector of mortality probabilities (the values recorded in any life table).

The survival curve ($S(t)$) is calculated as the cumulative product of $1 -$ mortality probability for each time period.

$$S(t) = \prod_{i=1}^{t} (1 - q_i)$$

where $q_i$ is the mortality probability at time $i$.

```{r gen_pop_surv_curve, fig.retina=100, fig.width=7, fig.height=5, fig.retina=2}
# Define a vector of mortality probabilities
df_life_table <- NeuroblastomaPSM::df_lifeTable_Jordan

# Calculate the survival curve
survival_curve <- NeuroblastomaPSM::get_lifeTable_survival_curve(
   mortality_probs = df_life_table$`Both sexes`
)

# Compare mortality probabilities to cumulative survival
rbind(
  head(cbind(
    survival_curve, 
    df_life_table$`Both sexes`
  )),
  tail(cbind(
    survival_curve,
    df_life_table$`Both sexes`
  ))
)

# Plot the survival curve obtained from the mortality probabilities
plot(
  x = df_life_table$`Age Group`,
  y = survival_curve, 
  type = 'l',
  xlab = "Age",
  ylab = "Survival",
  main = "Survival Curve from the Jordanian Life Table" 
)
```

### Adjust the mortatlity hazards:

The package provides the `interpolate_survival_curve` function which interpolate the hazard rates between the ages and apply user-defined multiplier. This function performs the following steps:

1. **Estimate Cumulative Hazards**:
   - The cumulative hazard at each age is estimated from the survival probabilities using the formula:
     $$H(t) = -\log(S(t))$$

2. **Fit Interpolation Model**:
   - A model is fitted to the log cumulative hazards as a function of age using either spline interpolation (`"spline"`) or a generalized additive model (`"gam"`). The choice of model is controlled by the `interpolation_model` parameter.

3. **Predict Cumulative Hazard Rates**:
   - Using the fitted model, the log cumulative hazards are predicted for the `target_ages`.
   - The predicted log cumulative hazards are then exponentiated to obtain the cumulative hazard rates:
     $$H(t) = \exp(\text{log_H}(t))$$

4. **Adjust Hazard Rates**:
   - The predicted cumulative hazard rates are adjusted by the `smr_multiplier`.

5. **Convert to Survival Curve**:
   - The adjusted cumulative hazard rates are converted back to survival probabilities using:
     $$S(t) = \exp(-H(t))$$

```{r adjust_surv_curve, fig.retina=100, fig.width=7, fig.height=5, fig.retina=2}
# Define age range and survival probabilities
curve_starting_age <- NeuroblastomaPSM::l_psm_parameters$age + NeuroblastomaPSM::l_psm_parameters$cure_threshold 
v_life_table_ages  <- df_life_table$`Age Group`
v_life_table_porbs <- df_life_table$`Both sexes`
v_life_table_porbs <- v_life_table_porbs[v_life_table_ages >= curve_starting_age]
v_life_table_ages  <- v_life_table_ages[v_life_table_ages >= curve_starting_age]

# Calculate the survival curve for the modeled cohort
survival_curve <- NeuroblastomaPSM::get_lifeTable_survival_curve(
   mortality_probs = v_life_table_porbs
)

# Plot the background survival curve for modeled cohort
plot(
  x = v_life_table_ages,
  y = survival_curve, 
  type = 'l',
  xlab = "Age",
  ylab = "Survival",
  main = "Survival Curve for 14-years Old" 
)

# Define output ages
target_ages <- seq(curve_starting_age, max(v_life_table_ages), 7/365)

# Interpolate Survival Curve
set.seed(1)
v_interpolated_curve <- NeuroblastomaPSM::interpolate_survival_curve(
  survival_curve_ages = v_life_table_ages,
  survival_curve = survival_curve,
  target_ages = target_ages
)

# Compare fitted and observed survival curves
plot(
  x = v_life_table_ages, 
  y = survival_curve, 
  xlab = "Age", 
  ylab = "Survival", 
  type = 'l',
  main = "Observed vs Fitted Survival Curves for 14-years Old" 
  )
lines(x = target_ages, y = v_interpolated_curve, col = "blue")
legend(
  "bottomleft", 
  legend = c("Observed survival curve", "Interpolated survival curve"),
  col = c("black", "blue"),
  lty = 1,
  bty = "n"
)
# Interpolate Survival Curve with a SMR = 5.6
set.seed(1)
v_interpolated_EFS_curve <- NeuroblastomaPSM::interpolate_survival_curve(
  survival_curve_ages = v_life_table_ages,
  survival_curve = survival_curve,
  target_ages = target_ages,
  smr_multiplier = NeuroblastomaPSM::l_psm_parameters$smr_EFS
)

# Compare fitted and observed survival curves
plot(
  x = v_life_table_ages, 
  y = survival_curve, 
  xlab = "Age",
  ylab = "Survival",
  type = 'l',
  main = "Observed, Fitted and EFS Survival Curves for 14-years Old" 
)
lines(x = target_ages, y = v_interpolated_curve,     col = "blue")
lines(x = target_ages, y = v_interpolated_EFS_curve, col = "orange")
legend(
  "bottomleft", 
  legend = c("Observed survival curve", "Interpolated survival curve", "EFS curve"),
  col = c("black", "blue", "orange"),
  lty = 1,
  bty = "n"
)
# Interpolate Survival Curve with a SMR = 5.6 * 1.9
set.seed(1)
v_interpolated_PPS_curve <- NeuroblastomaPSM::interpolate_survival_curve(
  survival_curve_ages = v_life_table_ages,
  survival_curve = survival_curve,
  target_ages = target_ages,
  smr_multiplier = NeuroblastomaPSM::l_psm_parameters$smr_PPS
)

# Compare fitted and observed survival curves
plot(
  x = v_life_table_ages,
  y = survival_curve,
  xlab = "Age",
  ylab = "Survival",
  type = 'l',
  main = "Observed, Fitted, EFS and PPS Survival Curves for 14-years Old" 
)
lines(x = target_ages, y = v_interpolated_curve,     col = "blue")
lines(x = target_ages, y = v_interpolated_EFS_curve, col = "orange")
lines(x = target_ages, y = v_interpolated_PPS_curve, col = "red")
legend(
  "bottomleft", 
  legend = c("Observed survival curve", "Interpolated survival curve", "EFS curve", "PFS curve"),
  col = c("black", "blue", "orange", "red"),
  lty = 1,
  bty = "n"
)
```

## Employ the 'cure threshold':

The 'cure threshold' assumptions in this PSM are implemented in the `calculate_markov_trace` function. Hence, part of the this function code is shown below to demonstrate how the 'cure threshold' assumptions are applied.

```{r cure_threshold, fig.retina=100, fig.width=7, fig.height=5, fig.retina=2}
# Define age range and survival probabilities
model_starting_age <- NeuroblastomaPSM::l_psm_parameters$age
curve_starting_age <- NeuroblastomaPSM::l_psm_parameters$age + NeuroblastomaPSM::l_psm_parameters$cure_threshold 
v_life_table_ages  <- df_life_table$`Age Group`
v_life_table_porbs <- df_life_table$`Both sexes`
v_life_table_porbs <- v_life_table_porbs[v_life_table_ages >= curve_starting_age]
v_life_table_ages  <- v_life_table_ages[v_life_table_ages >= curve_starting_age]

# Load the fitted Gompertz model and other parameters
models_fit <- NeuroblastomaPSM::parametric_models

# Define model parameters
l_params <- c(
  time_horizon = 15,
  cycle_length = 1/12,
  disc_rate_costs = 0.035,
  disc_rate_qalys = 0.015,
  NeuroblastomaPSM::l_psm_parameters
)

# Predict cumulative survival from the fitted parametric survival curves
df_survival_curves_long <- NeuroblastomaPSM::predict_cumulative_survival(
  models_fit = models_fit,
  l_params = l_params
)

# Flip survival dataframe to wide format:
df_survival_curves <- stats::reshape(
  data = df_survival_curves_long,
  timevar = "end_point",
  idvar = c("time", "treatment"),
  direction = "wide"
)

# Renaming the columns since reshape() adds a prefix to the column name:
names(df_survival_curves) <- gsub(
  pattern = "survival\\.",
  replacement = "",
  x = names(df_survival_curves)
)

# Calculating 'PPS' and 'D' state occupancy, no "cure threshold" assumptions:
df_survival_curves$PPS <- df_survival_curves$OS - df_survival_curves$EFS
df_survival_curves$D <- 1 - df_survival_curves$OS
df_survival_curves2 <- df_survival_curves

# Calculating 'PPS' and 'D' state occupancy, "cure threshold" assumptions:
if(l_params$time_horizon > l_params$cure_threshold) {
  # Get parameters
  model_starting_age  <- l_params$age
  curve_starting_age  <- model_starting_age + l_params$cure_threshold
  curve_end_age       <- model_starting_age + l_params$time_horizon
  cycle_length        <- l_params$cycle_length
  cure_threshold      <- l_params$cure_threshold
  cure_starting_cycle <- (cure_threshold / cycle_length) + 1
  v_curve_interpolation_ages <- seq(
    curve_starting_age + cycle_length,
    curve_end_age,
    cycle_length
  )
  
  # Prepare life table parameters
  v_life_table_ages  <- df_life_table$`Age Group`
  v_life_table_porbs <- df_life_table$`Both sexes`
  v_life_table_porbs <- v_life_table_porbs[
    v_life_table_ages >= curve_starting_age
  ]
  v_life_table_ages  <- v_life_table_ages[
    v_life_table_ages >= curve_starting_age
  ]
  
  # Calculate the survival curve for a healthy general population same age cohort
  v_lifeTable_survival_curve <- NeuroblastomaPSM::get_lifeTable_survival_curve(
   mortality_probs = v_life_table_porbs
  )
  # Estimate all-cause mortality for the EFS
  v_cure_EFS_curve <- NeuroblastomaPSM::interpolate_survival_curve(
    survival_curve_ages = v_life_table_ages,
    survival_curve = v_lifeTable_survival_curve,
    target_ages = v_curve_interpolation_ages,
    smr_multiplier = l_params$smr_EFS
  )
  # Estimate all-cause mortality for the PPS
  v_cure_PPS_curve <- NeuroblastomaPSM::interpolate_survival_curve(
    survival_curve_ages = v_life_table_ages,
    survival_curve = v_lifeTable_survival_curve,
    target_ages = v_curve_interpolation_ages,
    smr_multiplier = l_params$smr_PPS
  )
  # Apply the cure threshold assumptions
  df_GD2 <- df_survival_curves[df_survival_curves$treatment == "GD2", ]
  df_GD2$EFS[cure_starting_cycle:nrow(df_GD2)] <- cumprod(
    c(df_GD2$EFS[cure_starting_cycle], v_cure_EFS_curve)
  )
  df_GD2$PPS[cure_starting_cycle:nrow(df_GD2)] <- cumprod(
    c(df_GD2$PPS[cure_starting_cycle], v_cure_PPS_curve)
  )
  df_GD2$D <- 1 - (df_GD2$EFS + df_GD2$PPS)
  
  df_TT  <- df_survival_curves[df_survival_curves$treatment == "TT", ]
  df_TT$EFS[cure_starting_cycle:nrow(df_TT)] <- cumprod(
    c(df_TT$EFS[cure_starting_cycle], v_cure_EFS_curve)
  )
  df_TT$PPS[cure_starting_cycle:nrow(df_TT)] <- cumprod(
    c(df_TT$PPS[cure_starting_cycle], v_cure_PPS_curve)
  )
  df_TT$D <- 1 - (df_TT$EFS + df_TT$PPS)

  df_survival_curves <- rbind(df_GD2, df_TT)
}

# Extract relevant columns:
df_markov_trace <- df_survival_curves[
  ,
  c("time", "treatment", "EFS", "PPS", "D")
]
df_markov_trace2 <- df_survival_curves2[
  ,
  c("time", "treatment", "EFS", "PPS", "D")
]

# Sanity check:
stopifnot(
  "Markov trace does not sum up to 1." =
    all(rowSums(df_markov_trace[, c("EFS", "PPS", "D")]) == 1.00000)
)
```

The plots below show the difference between employing and not making the 'cure threshold' assumptions.

```{r ploting_cure_assumptions, fig.retina=100, fig.width=7, fig.height=5, fig.retina=2}
# Plot state occupancy - Cure Threshold
## TT 
plot(
  x = df_markov_trace$time[df_survival_curves$treatment == "TT"], 
  y = df_markov_trace$EFS[df_survival_curves$treatment == "TT"],
  ylim = c(0, 1),
  col = "darkgreen",
  xlab = "Age",
  ylab = "Survival",
  type = 'l',
  main = "TT State Occupancy - Cure Threshold" 
)
lines(
  x = df_markov_trace$time[df_survival_curves$treatment == "TT"], 
  y = df_markov_trace$PPS[df_survival_curves$treatment == "TT"],
  col = "orange"
)
lines(
  x = df_markov_trace$time[df_survival_curves$treatment == "TT"], 
  y = df_markov_trace$D[df_survival_curves$treatment == "TT"],
  col = "red"
)
legend(
  "topright", 
  legend = c("EFS", "PPS", "Dead"),
  col = c("darkgreen", "orange", "red"),
  lty = 1,
  bty = "n"
)

## GD2
plot(
  x = df_markov_trace$time[df_survival_curves$treatment == "GD2"], 
  y = df_markov_trace$EFS[df_survival_curves$treatment == "GD2"],
  ylim = c(0, 1),
  col = "darkgreen",
  xlab = "Age",
  ylab = "Survival",
  type = 'l',
  main = "GD2 State Occupancy - Cure Threshold" 
)
lines(
  x = df_markov_trace$time[df_survival_curves$treatment == "GD2"], 
  y = df_markov_trace$PPS[df_survival_curves$treatment == "GD2"],
  col = "orange"
)
lines(
  x = df_markov_trace$time[df_survival_curves$treatment == "GD2"], 
  y = df_markov_trace$D[df_survival_curves$treatment == "GD2"],
  col = "red"
)
legend(
  "topright", 
  legend = c("EFS", "PPS", "Dead"),
  col = c("darkgreen", "orange", "red"),
  lty = 1,
  bty = "n"
)
```

```{r ploting_no_cure_assumptions, fig.retina=100, fig.width=7, fig.height=5, fig.retina=2}
# Plot state occupancy - No Cure Threshold
## TT 
plot(
  x = df_markov_trace2$time[df_survival_curves$treatment == "TT"], 
  y = df_markov_trace2$EFS[df_survival_curves$treatment == "TT"],
  ylim = c(0, 1),
  col = "darkgreen",
  xlab = "Age",
  ylab = "Survival",
  type = 'l',
  main = "TT State Occupancy - No Cure Threshold" 
)
lines(
  x = df_markov_trace2$time[df_survival_curves$treatment == "TT"], 
  y = df_markov_trace2$PPS[df_survival_curves$treatment == "TT"],
  col = "orange"
)
lines(
  x = df_markov_trace2$time[df_survival_curves$treatment == "TT"], 
  y = df_markov_trace2$D[df_survival_curves$treatment == "TT"],
  col = "red"
)
legend(
  "topright", 
  legend = c("EFS", "PPS", "Dead"),
  col = c("darkgreen", "orange", "red"),
  lty = 1,
  bty = "n"
)

## GD2
plot(
  x = df_markov_trace2$time[df_survival_curves$treatment == "GD2"], 
  y = df_markov_trace2$EFS[df_survival_curves$treatment == "GD2"],
  ylim = c(0, 1),
  col = "darkgreen",
  xlab = "Age",
  ylab = "Survival",
  type = 'l',
  main = "GD2 State Occupancy - No Cure Threshold" 
)
lines(
  x = df_markov_trace2$time[df_survival_curves$treatment == "GD2"], 
  y = df_markov_trace2$PPS[df_survival_curves$treatment == "GD2"],
  col = "orange"
)
lines(
  x = df_markov_trace2$time[df_survival_curves$treatment == "GD2"], 
  y = df_markov_trace2$D[df_survival_curves$treatment == "GD2"],
  col = "red"
)
legend(
  "topright", 
  legend = c("EFS", "PPS", "Dead"),
  col = c("darkgreen", "orange", "red"),
  lty = 1,
  bty = "n"
)
```
