---
title: "Partition Survival Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Partition Survival Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  results = 'markup',
  message = FALSE, 
  warning = FALSE,
  comment = "#>",
  fig.width = 8, 
  fig.height = 6
)
```

## Introduction:

A partitioned survival model (PSM) with N distinct states quantifies the likelihood of a patient occupying each of these states at any specific time, given a certain treatment. This model relies on overlapping survival curves to estimate state membership, requiring a total of N-1 such curves for an N-state configuration.

The model uses the cumulative survival function, $S_n(t)$, to define the chance of a patient either reaching or surpassing a certain health state $n$ after time $t$. For the initial state, the probability is directly indicated by $S_1(t)$. The determination of membership for intermediate states, from the second to the $(n-1)^{th}$, is achieved through the difference $S_{n}(t) - S_{n-1}(t)$. The likelihood that a patient advances to the terminal state $N$, often signifying death, is given by $1-S_{N-1}(t)$, essentially the complement of the cumulative survival up to the next-to-last state.

For each treatment strategy $k$, the model predicts separate survival curves over time $t$ for each treatment approach $k$ and individual patient $i$, based on $L$ parameters $\alpha_l$, which may be modified by covariates $x_l$,

$$
\begin{aligned}
S_{n}(t|\alpha_{1n}(x_{1n, ik}), \ldots, \alpha_{Ln}(x_{Ln, ik})).
\end{aligned}
$$

Quality-adjusted life years (QALYs) and total costs for each health state, under a specific treatment and patient scenario, are derived by integrating the survival probability over time, adjusted by quality of life weights and costs. For any health state $h$, the calculations for QALYs and costs in the $m^{th}$ costs category are represented as:

$$
\begin{aligned}
\rm{QALYs}_{hik} &= \int_{0}^{T} q_{hik}(t) e^{-rt} p_{hik}(t)dt \\
\rm{Costs}_{m,hik} &= \int_{0}^{T} c_{m,hik}(t) e^{-rt} p_{hik}(t)dt
\end{aligned}
$$

Here, $q_{hik}$ denotes the quality-of-life weight, $c_{m,hik}$ represents annualized costs, $r$ stands for the discount rate, and $p_{hik}(t)$ signifies the probability of being in a specific health state at time $t$. These calculations span over $T$ time periods to account for QALYs and associated costs.

## The model:

### Model parameters:

The model parameters include the time horizon of decision problem and states payoffs or rewards.

```{r parameters}
# Model parameters
params <- list()
params$time_horizon <- 10
params$cycles_per_year <- 12
params$c_TT  <- 2000 # place holder
params$c_GD2 <- 3000 # place holder
params$c_PPS <- 3200 # place holder
params$u_EFS <- 0.23 # place holder
params$u_PPS <- 0.23 # place holder
params$disc_rate <- 0.035
```

### Survival curves:

The analysis documented by the **Survival functions** vignette suggests that **Gompertz** model fits the data best. The parameters of the fitted *Gompertz* distribution are employed in the section below to predict the cumulative survival curves over the model time horizon. 

```{r cumulative_survival}
# Load the fitted `Gompertz` model parameters
models_fit <- NeuroblastomaPSM::parametric_models

# Get the months from 0 to time horizon
time_points <- seq(
  from = 0, 
  to = params$time_horizon, 
  by = 1 / params$cycles_per_year
)

# Predict cumulative survival
df_survival_curves_long <- lapply(
  X = names(models_fit) |>
    `names<-`(names(models_fit)),
  FUN = function(curve_nm) {
    model_fit <- models_fit[[curve_nm]]
    
    end_point <- ifelse(
      test = grepl(pattern = "OS", x = curve_nm),
      yes = "OS",
      no = "EFS"
    )
    
    treatment_nm <- gsub(
      pattern = paste0(".", end_point),
      replacement = "",
      x = curve_nm
    )

    NeuroblastomaPSM::predict_survival_curve(
      surv_mod = model_fit,
      treatment_name = treatment_nm,
      end_point = end_point,
      time_points = time_points
    )
  }
) |>
  do.call(
    what = rbind,
    args = _
  ) |> 
  `rownames<-`(NULL)

# View predicted cumulative survival dataframe
rbind(
  head(df_survival_curves_long, n = 5), 
  tail(df_survival_curves_long, n = 5)
)
```

### State membership (Markov trace):

This model is represented by three health states, *event free survival (EFS)*,  *post progression survival (PPS)* and *death*. Using the predicted *EFS* and *OS*, state membership is estimated below.

```{r markov_trace}
# Pivoting wider using reshape() - long-formatted to wide-formatted
df_survival_curves <- reshape(
  data = df_survival_curves_long,
  timevar = "end_point",
  idvar = c("time", "treatment"),
  direction = "wide"
)

# Renaming the columns since reshape() adds a prefix to the column names
names(df_survival_curves) <- gsub("survival\\.", "", names(df_survival_curves))

# Creating new columns
df_survival_curves$PPS <- df_survival_curves$OS - df_survival_curves$EFS
df_survival_curves$D <- 1 - df_survival_curves$OS

# Selecting and ordering the desired columns (select in tidyverse)
df_survival_curves <- df_survival_curves[, c("time", "treatment", "EFS", "PPS", "D")]

# View wide formatted predicted cumulative survival dataframe
rbind(
  head(df_survival_curves, n = 5), 
  tail(df_survival_curves, n = 5)
)

# Markov trace
m_markov_trace <- df_survival_curves[, c("EFS", "PPS", "D")]

# Confirm markov trace is valid
stopifnot(
  "Markov trace does not sum up to 1." =
    all(rowSums(m_markov_trace) == 1)
)

# Visualize the markov trace
df_survival_curves |>
  reshape(
    data = _,
    varying = list(c("EFS", "PPS", "D")),
    v.names = "prop",
    timevar = "State",
    times = c("EFS", "PPS", "D"),
    direction = "long"
  ) |>
  ggplot2::ggplot(
    mapping = ggplot2::aes(
      x = time,
      y = prop,
      col = State
    )
  )+
  ggplot2::geom_step() +
  ggplot2::theme_classic() +
  ggplot2::scale_y_continuous(name = "Proportion") +
  ggplot2::scale_x_continuous(name = "Time (years)") +
  ggplot2::scale_color_manual(
    name = "State",
    values = c("D"= 1,
               "EFS" = 2,
               "PPS" = 3),
    labels = c("D"="Dead",
               "EFS" = "EFS",
               "PPS" = "PPS")
  ) +
  ggplot2::facet_wrap(~treatment)
```

### Economic Analysis:

```{r economic_analysis}
# create matrices for matrix multiplication
m_TR_Isotretinoin <- as.matrix(
  df_survival_curves[
    df_survival_curves$treatment == "Isotretinoin", c("EFS", "PPS", "D")
  ]
)
m_TR_Dinutuximab_β <- as.matrix(
  df_survival_curves[
    df_survival_curves$treatment == "Dinutuximab β", c("EFS", "PPS", "D")
  ]
)

# create vectors of health states payoffs
v_C_Isotretinoin <- c("EFS" = params$c_TT,
                      "PPS" = params$c_PPS,
                      "D" = 0)

v_C_Dinutuximab_β <- c("EFS" = params$c_GD2,
                       "PPS" = params$c_PPS,
                       "D" = 0)
v_Util <- c("EFS" = params$u_EFS,
         "PPS" = params$u_PPS,
         "D" = 0)

# Matrix multiplication of trace by:
## cost per cycle - gives total cost by cycle
v_costs_Isotretinoin  <- m_TR_Isotretinoin %*% 
  (v_C_Isotretinoin / params$cycles_per_year)
v_costs_Dinutuximab_β <- m_TR_Dinutuximab_β %*% 
  (v_C_Dinutuximab_β / params$cycles_per_year)

## utility by cycle - gives total utility by cycle
v_qalys_Isotretinoin  <- m_TR_Isotretinoin %*% 
  (v_Util / params$cycles_per_year)
v_qalys_Dinutuximab_β <- m_TR_Dinutuximab_β %*%
  (v_Util / params$cycles_per_year)

# Combine payoffs' calculations
m_results <- cbind(
  "c_Isotretinoin" = v_costs_Isotretinoin,
  "c_Dinutuximab_β" = v_costs_Dinutuximab_β,
  "q_Isotretinoin" = v_qalys_Isotretinoin,
  "q_Dinutuximab_β" = v_qalys_Dinutuximab_β
) |>
  `colnames<-`(c("c_Isotretinoin", "c_Dinutuximab_β", "q_Isotretinoin", "q_Dinutuximab_β"))

# Calculate discount weights
v_dw <- 1 /(1 + params$disc_rate) ^ time_points

# Apply discount weights to all columns of results matrix simultaneously
## this gives total discounted costs and qalys over the whole period.
m_disc_res <- v_dw %*% m_results |>
  `colnames<-`(c("c_Isotretinoin", "c_Dinutuximab_β", "q_Isotretinoin", "q_Dinutuximab_β"))

# Calculate incremental costs and utilities
inc_cost  <- m_disc_res[, "c_Dinutuximab_β"] - m_disc_res[, "c_Isotretinoin"]
inc_qaly  <- m_disc_res[, "q_Dinutuximab_β"] - m_disc_res[, "q_Isotretinoin"]

# calculate the ICER
as.numeric(inc_cost / inc_qaly)
```



